State-space PTA

### 2023-06-13 docs/manuscript/paper.tex
document updates

@@

\section{State-Space Model}
We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, $\bar{x}$, and the measurement state recorded by an observer,$\bar{y}$. We will take our state variable to be the intrinsic pulsar pulse frequency $f(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar. We will now consider how the intrinsic pulse frequency evolves in time, completely separate from the influence of a GW, and then go on to derive the  influence of a GW perturbation on the frequency recorded by an observer at Earth. \textcolor{red}{Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}

\subsection{Evolution of the pulsar frequency}
We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}. Within this model, $f$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering', achromatic variations in the pulse TOA intrinsic to the star). The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process. Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation,$\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g. a model of the neutron star crust and superfluid, components][]{Meyers2021}. For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.

\subsection{Modulation of pulsar frequency due to a GW}
In the presence of a GW, the $f(t)$  measured by an observer in the local NS reference frame is different from that measured by an observer on Earth. We want to determine quantitatively determine this difference - how the GW influences the received pulse frequency.

\subsubsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, \textcolor{red}{(TK: We have a notation conflict between $x$ = state and $x$ = coordinates)} where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources. Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{\textcolor{red}{TK: This point is important, since the phase offset is then the same between multiple pulsars. See also Melatos 2022 PT12}}. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
\begin{align}
	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
\end{align}
which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$
\begin{align}
	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}

\subsubsection{Pulse frequency as a photon}
\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	\nu = p_{\alpha} u^{\alpha}
\end{equation}
\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	f = p_t
\end{equation}
\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
\end{equation}
\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path. \footnote{\textcolor{red}{TK: See also e.g. Maggiore, Meltaos who takes the same approach...}} \newline

\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.

\subsubsection{Hamiltonian Mechanics}
The Hamiltonian in covariant notation can be written as 
\begin{equation}
	H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
\end{equation}
\noindent which if we substitute in our expression for the perturbed metric is
\begin{equation}
	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
\noindent Now, Hamilton's equations are
\begin{equation}
	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
\end{equation}

\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
\begin{equation}
	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
\end{equation}
\footnote{\textcolor{red}{TK: This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset}.}

\noindent Therefore the derivative w.r.t coordinate time $t$ is,
\begin{equation}
	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
\end{equation}
\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. It will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency. Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer \footnote{ \textcolor{red}{TK: see Melatos 2022, PT16}}. We will consider  the pulsar locations to be constant with respect to the Earth. In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB). This barycentering correction is typically applied during the pulsar detection process. Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar. The spatial coordinates along the photon path are related to the pulsar as $\bar{x}(t) = -\bar{q}t$. \newline

\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
\begin{equation}
	\dot{p}_t = A e^{i \gamma t + \Phi_0}
\end{equation}
where 
\begin{equation}
	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
\end{equation}
\footnote{\textcolor{red}{TK: compare with Melatos 2022, PT11}}
\noindent and
\begin{equation}
	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
\end{equation}

\subsubsection{Performing the integral}

The frequency shift experienced by the observer relative to the source due to a GW is then
\begin{align}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
	=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
	\label{eq:concise}
\end{align}
We can write this more explicitly as,
\begin{align} \label{eq:final}
	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
	&\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{align}
or alternatively, with $h_{ij} = g_{ij} - \eta_{ij}$,
\begin{equation}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
	\label{eq:freq_shift}
\end{equation}

\subsection{State-space structure} \label{sec3}
We now have a model for the evolution of the intrinsic pulsar frequency states and the measured frequency timeseries. We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
\begin{equation}
	df = \mathcal{A} f dt + N(t) dt + \sigma dB(t)
	\label{eq:state1}
\end{equation}
where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process). This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
\begin{equation}
	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
\end{equation} 
If we move from a solution in continuous time to discrete time then
\begin{equation}
	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
\end{equation}
where
\begin{equation}
	F_i = e^{A (t_{i+1} - t_i)}
\end{equation}
\begin{equation}
	T_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} N(t') dt'
\end{equation}
\begin{equation}
	\eta_i = \int_{t_i}^{t_{i+1}}  e^{A (t_{i+1} - t')} \sigma dt'
\end{equation}
Given the discrete solution $f(\bar{t})$ where $\bar{t} = (t_1, t_2, ...,t_N)$ the intrinsic frequency can be related to the measured frequency as 
\begin{equation}
	f_M (\bar{t})= f g(\bar{\theta},\bar{t}) + N_M
\end{equation}
where $g(\theta,t)$ ("measurement function") is a function of some constant parameters, $\bar{\theta}$, and discrete time $\bar{t}$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
\begin{equation}
	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
\end{equation}
for variance $\Sigma^2$. The measurement function follows from Eq. \ref{eq:freq_shift}
\begin{equation} \label{eq:final}
	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}
or in a more concise form where we have switched to a trigonometric form of the equations and made the time dependence explicit,
\begin{equation} 
	g(\bar{\theta},t) = 1 -A \cos(-\Omega t (1 + n\cdot q ) + \Phi_0)
\end{equation}
where 
\begin{equation}
	A =  \frac{1}{2} \frac{H_{ij} q^i q^j}{1 + \bar{n} \cdot \bar{q}}[1 - \cos\left( {\Omega(1+\bar{n} \cdot \bar{q}) d}\right)]
		\label{eq:state2}
\end{equation}

\section{Detection and parameter estimation}
Let's  review and categorise all the parameters of above state-space model. We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar, the GW parameters and the noise parameters i.e.
\begin{equation}
	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW} \cup \bar{\theta}_{\rm noise}
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm noise} = [\sigma, \Sigma]
\end{equation}
\textcolor{red}{TK: $\sigma$ is currently a parameter shared between all pulsars in current code implementation. Probably need to update this since different pulsars have different process noise, but need to check identifiability and likelihood curves - only enters the Kalman filter only through the process noise matrix...}. Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not. For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate. For now we will consider the measurement noise to be known, although in principle this too could be estimated. Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location. \newline

\noindent We want be able to infer the values of the model parameters and look for evidence of a GW signal in a data stream. To this end, we will use a Kalman filter in to obtain likelihood estimates of the data given a set of parameters. We will then deploy the Kalman filter in conjunction with a nested sampling technique for optimisation of the parameters over the likelihood. We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation. We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.

@@

@@

\subsection{Detection on synthetic data}\label{sec:detection}

>>>>

This is how we usually search for single source PTAs

estimate all unknown parameters optimally
here is a difference approach

Typical approaches for

@@

\section{State-Space Model}\label{sec:model}
We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer. For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar. We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.  In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively. We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}

\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}. Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering', achromatic variations in the pulse TOA intrinsic to the star). The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process. Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g. a model of the neutron star crust and superfluid, components][]{Meyers2021}. For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.

\subsection{Modulation of pulsar frequency due to a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local neutron star reference frame is different from that measured by an observer on Earth. We will now derive the influence of the GW on the received pulse frequency.

\subsubsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources. We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
		h_{\times} &= -2h\cos \iota
\end{align}
The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
\begin{align}
	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
\end{align}
which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coodinate $\phi$) and the polarisation angle $\psi$
\begin{align}
	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}

\subsubsection{Pulse frequency as a photon}
\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	f = p_{\alpha} u^{\alpha}
	\label{eq:freq_temporal}
\end{equation}
\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	f = p_t
\end{equation}
\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
	\label{eq:integral}
\end{equation}
\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.

\subsubsection{Hamiltonian Mechanics}
The Hamiltonian in covariant notation can be written as 
\begin{equation}
	H = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
\end{equation}
\noindent which if we substitute in our expression for the perturbed metric is
\begin{equation}
	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
\noindent Now, Hamilton's equations are
\begin{equation}
	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
\end{equation}

\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
\begin{equation}
	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
\end{equation}
\noindent The derivative w.r.t coordinate time $t$ is
\begin{equation}
	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
\end{equation}
\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. 
It will prove useful to recognise that 
\begin{equation}
	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
\end{equation} 
where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency. Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer. Similarly, we can parameterize the photon spatial coordinates $\bar{x}$ as,
\begin{equation}
	\bar{x}(t) = - \bar{q} (t - t_1)
\end{equation} 
Note that when our integration variable $t = t_1$ (i.e. the photon is at Earth) then $\bar{x}(t=t_1) = 0$ which is what we expect given our chosen coordinate system. Also note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. We will treat the pulsar locations to be constant with respect to the Earth, i.e. $\bar{q}$ is not a function of time. In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB). This barycentering correction is typically applied during the pulsar detection process. Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar. \newline

\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
\begin{equation}
	\dot{p}_t = A e^{i B t + C}
\end{equation}

with 
\begin{equation}
	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
\end{equation}
\begin{equation}
	B = -\Omega (1 + \bar{n}\cdot \bar{q}) 
\end{equation}
and 
\begin{equation}
	C = \Omega \tau \bar{n} \cdot \bar{q}  +\Phi_0
\end{equation}
From Eq \ref{eq:integral} the frequency shift experienced by the observer relative to the source due to a GW is then
\begin{align}
	&p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = \\
	 & A \int_{t = \tau - d}^{t=\tau} e^{i B t + C}dt  \\
	& =\frac{-iA}{B} e^{i(-\Omega \tau +\Phi_0)} \left(1 -e^{-i B d}\right) \\
	&= \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right)
	\label{eq:concise}
\end{align}
for $h_{ij} = g_{ij} - \eta_{ij}$. Since via Equation  \ref{eq:freq_temporal} we can related the temporal component of the covariant momentum to the frequency, it follows that
\begin{align}
	&f(\tau)|_{\rm Earth} = f(\tau - d)|_{\rm source}  \left( 1- \frac{\omega}{2} \frac{ h_{ij} (\tau; \bar{x}= 0)q^i q^j }{(1 + \bar{n}\cdot \bar{q}) }  \left(1 -e^{i \Omega (1 + \bar{n}\cdot \bar{q})  d}\right) \right)
	\label{eq:main_eq}
\end{align}
where we have identified $\omega$ with $f(\tau - d)|_{\rm source}$.

\section{Detection and parameter estimation}
Let's review and categorise all the free parameters,$\bar{\theta}$, of above model. We can generally separate these into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source
\begin{equation}
	\bar{\theta} =  \bar{\theta}_{\rm PSR} \cup \bar{\theta}_{\rm GW}
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm PSR} = [\gamma, f_{\rm EM}(0), \dot{f}_{\rm EM}(0),d]^{(N)}
\end{equation}
\begin{equation}
	\bar{\theta}_{\rm GW} = [h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0]
\end{equation}
Whilst the GW parameters are shared between measurements for each pulsar, the pulsar parameters are clearly not. For a PTA dataset of $N$ pulsars we have $7 + 5N$ parameters to estimate. Note that whilst this is a large parameter space, in general the pulsar parameters are much better constrained than the GW parameters: for example we have rough estimates for the pulsar distances accurate to $\sim$ 10$\%$, but we have no prior information on the source location. \newline

In this section we present a new method to infer the model parameters and calculate the marginal likelihood (i.e. the model evidence). In Section \ref{sec:statespace} we will present a discretised version of the model of Section \ref{sec:model} which maps onto the discretely sampled observable $f(\tau)|_{\rm Earth}$. In Section \ref{sec:kalman_filter} we will outline how observations of this pulse frequency can be used to recover the underlying intrinsic pulsar state $f(\tau -d)|_{\rm source}$ using a Kalman Filter. In Section \ref{sec:nested_sampling} we will demonstrate how to deploy the Kalman filter in conjunction with a nested sampling technique to recover the model parameters. We will now review the Kalman filter and the associated likelihood in Section \ref{sec:kalman_filter} before going on in Section \ref{sec:nested_sampling} to discuss how to use nested sampling methods together with the filter for detection (model selection) and parameter estimation.

\subsection{Discretised model} \label{sec:statespace}
We can express the intrinsic frequency evolution, Eq. \ref{eq:frequency_evolution_expanded}, in a alternative form as,
\begin{equation}
	df = \mathcal{A} f dt + N(t) dt + \sigma dB(t)
	\label{eq:state1}
\end{equation}
where $\mathcal{A} = -\gamma$, $N(t) = \gamma(f_{\rm EM}(0) + \dot{f}_{\rm EM}(0) t) +\dot{f}_{\rm EM}(0)$ and $dB(t)$ denotes increments of Brownian motion (Wiener process). This equation is easily identified as an Ornstein-Uhlenbeck process which has a general solution given by \citep{gardiner2009stochastic},
\begin{equation}
	f(t) = e^{\mathcal{A}t}f(0) + \int_0^t e^{\mathcal{A}(t-t')} N(t') dt' + \int_0^t e^{\mathcal{A}(t-t')} \sigma dB(t')
\end{equation} 
If we move from a solution in continuous time, $t$, to discrete time, $\bar{t} = (t_1, t_2, ...,t_K)$, then
\begin{equation}
	f(t_{i+1}) = F f(t_i) + T_i + \eta_i
\end{equation}
where
\begin{equation}
	F_i = e^{\mathcal{A} (t_{i+1} - t_i)}
\end{equation}
\begin{equation}
	T_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} N(t') dt'
\end{equation}
\begin{equation}
	\eta_i = \int_{t_i}^{t_{i+1}}  e^{\mathcal{A} (t_{i+1} - t')} \sigma dt'
\end{equation}
The discrete solution $f(\bar{t})$ to the intrinsic frequency can be related to the discrete measured frequency via Eq. \ref{eq:main_eq} as,
\begin{equation}
	f_M (\bar{t})= f(\bar{t}) g(\bar{\theta},\bar{t}) + N_M
\end{equation}
where $g(\theta,t)$ can be expressed in a trigonometric form as 
\begin{equation}
		X = 1 - \frac{1}{2} \frac{ H_{ij}q^i q^j }{(1 + \bar{n}\cdot \bar{q}) } \left[ \cos(-\Omega \tau +\Phi_0) - \cos(-\Omega \tau +\Phi_0 + \Omega (1 + \bar{n}\cdot \bar{q})  d) \right]
\end{equation} 
whilst $N_M$ is a Gaussian measurement noise that satisfies 
\begin{equation}
	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
\end{equation}
for variance $\Sigma^2$.

@@

\subsection{Practical considerations}

heterodyeing

pulsar terms 
For now we will consider the measurement noise to be known, although in principle this too could be estimated.

\section{Tests with synthetic data}

We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.

@@

\subsection{Detection}\label{sec:detection}
